Run AwAI: StarCraftII Adversary Avoidance Using Neural Networks and Genetic Algorithms

### 2018-06-04 doc/run_awai.tex
added sections

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{float}
\usepackage[margin=.75in]{geometry}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{multicol}

@@

\author{Kyle R. Chickering \\ krchicke@ucdavis.edu
  \and Alex R. Mirov \\ armirov@ucdavis.edu
  \and Simon H. Wu \\ simwu@ucdavis.edu
  \and Xin Jin \\ jin@ucdavis.edu}

@@

\begin{multicols}{2}
\section{Introduction}
Since AlphaGo's defeat of the Go champion Ke Jie, the artificial intelligence
community has
been looking for a new problem domain to apply learning techniques to. A
surprising candidate for artificial intelligence testing has
emerged in the form of real time strategy games. Real time strategy games work
by having a player make critical decisions in real time against an enemy
player. The game StarCraftII, released by Blizzard entertainment in 2010, is a
real time strategy
game that has captured the interest of the artificial intelligence community.
The game presents several challenges for artificial intelligence, the biggest of
which is the magnitude of possible game states. The game technically has a
finite number of game states, but the granularity of those states renders the
game practically continuous.

@@

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{chart}\label{fig:flow chart}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Control and data flow through the Run AwAI controller.}
\end{changemargin}
\end{figure}

@@

Our neural network uses the standard sigmoid function $\varsigma$ as its
activation function:
\begin{align*}
  \varsigma(x) = \frac{1}{1 + e^{-x}}
\end{align*}
We chose this function for a number of reasons, the least of which is its
popularity as an activation function in neural networks. Because of the shape of
the sigmoid function, inputs to subsequent neurons stay in the nice range of
$[-1, 1]$, which has the effect of normalizing the network. Since we are not
using backpropogation, the nice properties of the sigmoid function were
unimportant to us. However, if we were to do live backpropogation during
simulations (See section \ref{sec:improvements}), the sigmoid function would be
a useful activation function.

@@

\subsection{Controller}

@@

\begin{figure}[H]\label{fig:fitness graph}
\centering
\includegraphics[width=0.45\textwidth]{fig_1}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Graph of average fitness (blue) and top fitness (orange) for each
    generation in Trial 1. Top fitness is the fitness of the most fit individual
    at the end of each generation's simulations. Average fitness is calculated
    by averaging \eqref{eq:fitness} over all individuals $\bm{i}$ in the
    population. The breeding constant is set to 15, node generation rate is set
    to 0.3, new connections at 0.2 and non structural mutations at a rate of
    0.4}
\end{changemargin}
\end{figure}

@@

\begin{figure}[H]\label{fig:fitness graph}
\centering
\includegraphics[width=0.45\textwidth]{fig_2}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Graph of average fitness (blue) and top fitness (orange) for each
    generation in Trial 1. Top fitness is the fitness of the most fit individual
    at the end of each generation's simulations. Average fitness is calculated
    by averaging \eqref{eq:fitness} over all individuals $\bm{i}$ in the
    population. The breeding constant is set to 15, node generation rate is set
    to 0.5, new connections at 0.5 and non structural mutations at a rate of
    0.75}
\end{changemargin}
\end{figure}

@@

%% TODO: Talk about choice of hyper parameters, population size, episode count
%% mutation constants, eta in fitness function

\section{Conclusion}
Different selection types?
\blindtext

\subsection{Improvements}\label{sec:improvements}
% live back propagation
One of the for-most improvements we could do is to train the neural networks
during the simulation as well as training them through evolution. In addition
to improving the convergence, this strategy would mimic nature even more.
Organisms learn during their lifetimes, and there is speculation that the knowledge obtained by an individual during its lifetime can be passed down through the
genome.

@@

@@

\end{multicols}

@@

\begin{thebibliography}{9}
\bibitem{NEAT} 
  Kenneth O. Stanley, Risto Miikkulainen.
  \textit{Evolving Neural Networks through Augmenting Topologies}. 
  The MIT Press Journals.

@@

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{blindtext}
\usepackage{bm}
\usepackage{float}
\usepackage[margin=1.25in]{geometry}
\usepackage{graphicx}
\usepackage{mathrsfs}
\usepackage{multicol}

@@

\author{
    Kyle R. Chickering \\ {\footnotesize\textbf{krchicke@ucdavis.edu}}
    \and Alex R. Mirov \\ {\footnotesize\textbf{armirov@ucdavis.edu}}
    \and Simon H. Wu \\   {\footnotesize\textbf{simwu@ucdavis.edu}}
    \and Xin Jin \\       {\footnotesize\textbf{jin@ucdavis.edu}}
}

@@

%\begin{multicols}{2}
\section{Introduction}
Since AlphaGo's defeat of the Go champion Ke Jie, the artificial intelligence
community has
been looking for a new problem domain to apply learning techniques to. A
surprising candidate for artificial intelligence testing has
emerged in the form of real time strategy games. Real time strategy games work
by having a player make critical decisions in real time against an enemy
player. The game StarCraftII, released by Blizzard entertainment in 2010, is a
real time strategy
game that has captured the interest of the artificial intelligence community.
The game presents several challenges for artificial intelligence, the biggest of
which is the magnitude of possible game states. The game technically has a
finite number of game states, but the granularity of those states renders the
game practically continuous.

@@

\begin{figure}[h]\label{fig:flow chart}
\centering
\includegraphics[width=0.45\textwidth]{chart}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Control and data flow through the Run AwAI controller.}
\end{changemargin}
\end{figure}

@@

Our neural network uses the standard sigmoid function $\varsigma$ as its
activation function:
\begin{align*}
  \varsigma(x) = \frac{1}{1 + e^{-x}}
\end{align*}
We chose this function for a number of reasons, the least of which is its
popularity as an activation function in neural networks \cite{russell}. Because
of the shape of the sigmoid function, inputs to subsequent neurons stay in the
nice range of $[-1, 1]$, which has the effect of normalizing the network. Since
we are not using backpropogation, the nice properties of the sigmoid function
were unimportant to us. However, if we were to do live backpropogation during
simulations (See section \ref{sec:improvements}), the sigmoid function would be
a useful activation function.

@@

\subsection{Controller}
The controller was the interface we used to communicate with StarCraftII and run
the simulations need to commute the fitness function. Our controller has a
function to step the game simulation, which is called several times per
second. We implemented all of our agent's logic, consisting of feeding forward
the inputs through our neural network, inside of this step function.

We spent some time deciding what inputs to send to our neural network during the
simulation. Originally, we sent absolute coordinates, but we found that these
coordinates made the network into essentially a binary decision
network. Instead, we settled on giving the neural networks relative coordinates
produced much better results. We also give the neural network the relative
coordinates of the enemy units. The controller queries the PySC2 interface for
these values every time we take a step in the simulation, and the values are
pulled out of StarCraftII itself.

% The Pysc2 interface allowed us to create an agent to interact with the Star
% Craft II game.  The agent class contains a \texttt{step()} function that is
% called several times per second of game time. This meant that all of the actions
% the agent took had to be in this step function, and thus this is where we would
% integrate our neural net.  We decided to use the mini game map DefeatRoaches,
% which is included with Pysc2 as a base for our simulation environment. By
% heavily relying on a preexisting map and mini game, we could focus more of our
% time and resources on developing the genetic algorithm and neural network
% implementation.

% To make our controller implementation as modular as possible, we decided to
% provide a dictionary of inputs and receive a dictionary of outputs. The inputs
% would be the x and y coordinates of the agent’s units, as well as the x and y
% coordinates of the enemy units. The inputs had to be created and assembled into
% a dictionary in the step function.  In order to normalize the inputs, we divided
% the coordinate value by the maximum map size in the corresponding axis.  To
% obtain the values for the inputs, we created a series of methods for our agent
% class.  These methods include \texttt{getCurrentLocation()} and
% \texttt{getCurrentEnemeyLocation}().  We also created methods to move the
% friendly units, including \texttt{moveToLocation()} and setTargetDestination.
% These functions could be used together to set a target destination coordinate
% and cause the friendly units to move toward that target.

One problem we encountered that we had to correct was determining when units
arrived at their target destination.  We solved this by comparing the target
destination and the unit's current location every time we step the simulation.
Because the mini game uses multiple units (both friendly and hostile), the unit
location coordinates are averaged over all selected units locations.  This means
that often units will never actually arrive at their precise destination, just
get very close to it. To mitigate this issue, we decided to consider units
within a certain radius of their target destination as at their destination.
Once this condition is met, we could ask the simulation whether or not units had
reached their destination. Once the units reach their destination, we can choose
a new destination for them to move to.

% Whenever \texttt{step()} is called by Pysc2, the first thing we do is check the
% observable action space. If friendly unit movement is not an option, this means
% the unit(s) are not selected.  If units are not selected, this means it is the
% first time the \texttt{step()} function is being called in the simulation
% episode.  In this case, in addition to selecting the units, we also reset
% various counters (such as steps survived).  During the first call of
% \texttt{step()}, we also unpickle a neural network file.

The controller also handles the interpretation of the data from the PySC2 interface, and acts as a ``human'' playing the game. The controller is responsible for selecting movements, making decisions, and executing those decisions.

% In the case where \texttt{step()} is called and units are already selected, the
% \texttt{moveToLocation()} function is called.  Initially, we set the target
% destination to the enemy location.  We have our units move directly to the
% hostile units in order to engage them in combat, and this initiate a pursuit
% from the enemy.  We felt this would be the most direct way to initiate a
% pursuit.  We investigated running two Pysc2 agents against each other, however
% Pysc2 currently only support the simulation of one custom agent at a time. Once
% the friendly units arrive at their destination, the \texttt{moveToLocation()}
% function returns an arrived at target status to the calling \texttt{step()}
% function.

When the controller receives an arrived at target status, a new
target destination is set.  To generate a new target destination, the relevant
aspects of the current state (current location of friendly and enemy units) are
captured and fed into the neural network.  The neural network returns a
vector of the action space and a value between -1 and 1 for each
action label key. The action with the highest number will be chosen, and then
taken.

Our action space is movements that the friendly units can take. We decided to
quantize the action space into discrete movements, as opposed to a continuous
$x$-$y$ coordinate plane. By only allowing our units to move in the 4 cardinal
directions, diagonals, or stay put, we hoped our first implementation of the
neural network and genetic algorithm would have an easier time converging.

%% Fix
To evaluate the fitness of the neural network, we count the number of steps that
the units survived for.  One problem we encountered and solved, was the initial
series of steps taken before engaging the enemy units. For example, the
DefeatRoaches map spawns the enemy and friendly units on opposite sides of the
map, in a random selection of several predefined start locations.  However, the
distance between these start locations can vary between simulation episode.
This means our neural network fitness (number of steps survived) could be
slightly affected by an irrelevant condition.  To mitigate this issue, we
subtract the number of steps it takes until the friendly units reach their first
destination (the location of the enemy units) and start their engagement. By
only counting steps after enemy engagement has begun, our fitness function is
more reflective of survival time.

% In order for our evolutionary genetic algorithm to evolve, the end fitness of a
% network must be returned from the Pysc2 simulation.  Our simulation runs a
% configurable (up to 55) episodes per individual (neural network) and averages
% the fitness for all of those simulation episodes. We played with the number of
% simulation episodes per simulation test, exploring the tradeoff between
% simulation runtime and precision.  In the future, we could explore modifying our
% fitness function to incorporate more meaning from having multiple simulation
% episodes.  One future idea we had was to see if the group of individual survival
% times were skewed above or below the mean.

In addition, the controller is responsible for calculating $\mathcal{S}(e)$ for
each episode (simulation) that it runs, and returning the value to our genetic
algorithm. This allows us to make modifications to our fitness function without
having to make changes to the genetic algorithm. This also gives us the
flexibility to extend the capabilities of the controller to run different style
simulations in the future. For example we could run the agent through multiple
types of maps, and compute the average fitness between multiple different
episodes and maps.

The average fitness for the individual is then saved in a Python pickle file,
along with the fitness of the top (most fit) individual.  Because the Pysc2
simulation is invoked from the genetic algorithm code, the algorithm can then
unpickle the file.  After all individuals in a generation are tested and their
fitnesses collected, the genetic algorithm can select the most fit individuals
(neural networks) for breeding.

@@

\begin{figure}[h]\label{fig:fitness graph}
\centering
\includegraphics[width=0.45\textwidth]{fig_1}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Graph of average fitness (blue) and top fitness (orange) for each
    generation in Trial 1. Top fitness is the fitness of the most fit individual
    at the end of each generation's simulations. Average fitness is calculated
    by averaging \eqref{eq:fitness} over all individuals $\bm{i}$ in the
    population. The breeding constant is set to 15, node generation rate is set
    to 0.3, new connections at 0.2 and non structural mutations at a rate of
    0.4}
\end{changemargin}
\end{figure}

@@

\begin{figure}[h]\label{fig:fitness graph}
\centering
\includegraphics[width=0.45\textwidth]{fig_2}
\begin{changemargin}{0.15in}{0.15in}
  \caption{Graph of average fitness (blue) and top fitness (orange) for each
    generation in Trial 1. Top fitness is the fitness of the most fit individual
    at the end of each generation's simulations. Average fitness is calculated
    by averaging \eqref{eq:fitness} over all individuals $\bm{i}$ in the
    population. The breeding constant is set to 15, node generation rate is set
    to 0.5, new connections at 0.5 and non structural mutations at a rate of
    0.75}
\end{changemargin}
\end{figure}

@@

\section{Conclusion}
Different selection types?

\subsection{Improvements}\label{sec:improvements}
One of the foremost improvements we could do is to train the neural networks
during the simulation as well as training them through evolution. In addition
to improving the convergence, this strategy would mimic nature even more.
Organisms learn during their lifetimes, and there is speculation that the knowledge obtained by an individual during its lifetime can be passed down through the
genome.

@@

In the future, we could expand the output action space to a continuous $x$ and
$y$ coordinate plane, as opposed to discrete directions. Although a larger
action space can often require more iterations of the genetic algorithm to
converge, the benefits would be more nuanced agent behavior, and more realistic
agent movements. As it stands now, our agent moves in what may be considered a
very unnatural way, and while this does not affect the agent's ability to
preform, it could be seen as a disadvantage in certain situations where the fact
that the game is being played by an AI may be undesirable.

We have also talked about putting more attention towards our fitness function. An interesting possibility is to take into account the number of episodes in $\bm{\mathcal{E}}$, and rewarding agents that have more simulations. We can then associate individuals with their simulation meta-data, and save time running simulations. We can start with a low number of simulations per individual, and as the generations progress, increase the number of simulations per individual. This would have the effect of starting with relatively low fitness, but increasing the ``accuracy'' of the fitness over time.

@@

%\end{multicols}

@@

\begin{thebibliography}{9}

@@

\bibitem{NEAT} 
  Kenneth O. Stanley, Risto Miikkulainen.
  \textit{Evolving Neural Networks through Augmenting Topologies}. 
  The MIT Press Journals.
